### 2020-02-12 paper/sex_selection_spacing-ver3.tex
Version close to submit worthy
Just over 8,000 words.

\usepackage{endfloat}

@@

\date{January 2019}

@@

% Demography abstract
\noindent

@@

@@

Understanding how birth spacing changes is of interest for two main reasons, in 
addition to a better general understanding how fertility decisions are made.

@@

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility behavior taking into
account both the likelihood of parity progression and having a child of a specific sex.

% [Hypotheses / questions]
There are four main set of results.

First, birth intervals have increased for all education groups over the four decades, 
with larger increases, the higher the parity and the higher the education level.
Women who are most likely to use sex selection---well-educated women with no
sons---have seen the most substantial increases in birth intervals.
[TK check this] As a result, well-educated women now show a reversal of the traditional spacing 
pattern, with the longest intervals for women with no sons.
Those least likely to use sex selection, women with less education in rural areas, 
still follow the traditional pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
[TK add sex ratios and parity progression discussions]

Second, there is little difference over time and across groups in the likelihood of 
having a very short birth interval given the sex composition of prior children, 
especially for the lower parities.
Hence, the increases in average spacing come predominantly from the longer birth intervals
getting even longer over time, driven to a large degree by a substantial use of sex 
selection.
The exception to this result is that the better-educated women with \emph{very high} 
predicted use of sex selection appear to conceive later exactly because the next child is 
almost certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell 
in the 1990s and early 2000s, when spacing began to increase as the result of secular 
changes and increasing use of sex selection.
Although predicted cohort fertility and the fertility rate has lately begun to converge,
the predicted cohort fertility is still ten to twenty percent higher than the
fertility rate.
Furthermore, even with the convergence, predicted cohort fertility is still at or above 
replacement level, except for the best-educated urban women.

Finally, infant mortality risk has decline substantially over time for all groups,
but fastest for the lower education groups, who are now close to the level of women with 
the most education.
Mortality risk is, however, still inversely related to education level, especially for 
very short birth intervals.
There is no evidence that the increasing use of sex selection is associated with higher 
mortality risk.

@@

The NFHS data show a U-shaped relationship between education and working for married
women, with the highest percent working for women with either no education 
or with twelve or more years and the lowest for women with eight to eleven years of 
education.%
\footnote{
See Appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}
Women are more likely to report working if they live in rural than urban areas 
and the older they are.
For married women there is relatively little change in the percentage currently working 
over time,
but all education and age groups have become substantially more likely to work for
a family member.

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the rapid expansion in female education, and the 
associated change in the composition of better-educated women, is that the behavior
of the better-educated would change. 
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, the women adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation is a first indication that this
process still operates.

@@

To understand the potential implications for birth spacing of the substantial changes in 
income, female education, and female labor force participation that India has experienced, 
this section discuss relevant theories and prior findings on birth spacing.

The standard economic argument for why increasing female education has predominantly
been associated with shorter spacing is that parents incur time costs when they have 
children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

The low and declining female labor force participation, especially for younger women, 
suggests, however, that families face little incentive to space children more closely 
together for economic reasons.
This may happen if household income has increased so substantially that the income effect 
dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70 percent higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appear to be more negatively elastic to their husbands' 
wages than it is positively elastic to own wages \citep{Bhargava2018}.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than the standard income effect alone predicts.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing potential income for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.
If more and ``better'' parental attention per child results in higher child ``quality'' 
we should expect longer birth intervals 
\citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is one crucial aspect of child quality---health and mortality---where
longer spacing does lead to better outcomes, although this relationship weakens with 
maternal education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

The introduction of sex selection allows parents to avoid the birth of girls,
but increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility, and, for a 
given desired number of children, the higher the parity \citep{Portner2015b}.
The increased use of sex selection in India with education and in urban
areas is consistent with a lower desired fertility for both groups
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated women also tend to live in households with higher income, which better 
enables them to access prenatal sex determination and lowers the relative costs of using 
sex selection and have long birth intervals.

We may even observe a ``delay'' effect, where parents conceive later because they know 
that short spacing is detrimental to the next child's health, and that the next child is 
more likely to be a boy because of access to sex selection.
Working in the opposite direction is that women with more education can, in principle, 
space their children closer together without substantially increasing child mortality 
risk.
Given the apparent lack of pressure to return to the labor force,
this effect is, however, likely to be small.

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
As desired fertility decreases with increasing education, I, furthermore, expect 
birth spacing to increase the most among the better-educated because of the larger
increase in household income income---even with declining female labor participation---and 
their use of sex selection.
Even with the large increase in the number of better-educated women, "Sanskritization" 
imply that the changing composition does not substantially change this group's use of 
sex selection.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.

The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.

@@

The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from nine months after one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.%
\footnote{
I discuss the choice of which variables to use for non-proportionality in more detail
in the ``Explanatory Variables'' section.
}

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is an interval---here equal
to three months---the model can be estimated using a standard multinomial logit model.
%
% \footnote{
% The multinomial model assumes that alternative exit states are stochastically independent
% (Independence of Irrelevant Alternatives or IIA), which rules out any individual-specific 
% unmeasured or unobservable factors that affect both the hazard of having a girl and the 
% hazard of having a boy.
% I, therefore, include a proxy for fecundity discussed in Section \ref{sec:data}.
% Also, the multivariate probit model can be used as an alternative because the IIA is not 
% imposed \citep{han90}.
% The results are substantially identical between these two models and
% available upon request.
% }
In the reorganized data the outcome variable is 0 if the woman does not have a child in a 
given interval (the base outcome), 1 if she gives birth to a son in that interval, and 2 
if she gives birth to a daughter in that interval.

@@

Interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy, $b$, and of having a girl, $g$, in period $t$, conditional on 
a set of explanatory variables and not having had a child before that period.
% , as
% \begin{align}
% P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
% {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})}
% \label{eq:probability_boy} \\
% P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
% {1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}
% \label{eq:probability_girl}
% \end{align}
It is then straightforward to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent boys born over the entire spell.

@@

The data come from the four rounds of the National Family Health Survey
collected in 1992--1993, 1998--1999, 2005--2006, and 2015--2016.%
\footnote{
A delay in the survey for Tripura means that NFHS-2 has some observation 
collected in 2000.
}
The surveys are large: NFHS-1 covered 89,777 ever-married women 
aged 13--49 from 88,562 households;
NFHS-2 covered 90,303 ever-married women aged 15--49 from 92,486 households;
NFHS-3 covered 124,385 never-married and ever-married women aged 
15--49 from 109,041 households;
and 
NFHS-4 covered 699,686 never-married and ever-married women aged
15--49 from 601,509 households surveyed.

I exclude visitors to the household, as well as women who fall into any of the following 
categories: never married; no gauna yet; married more than once; divorced; 
not living with husband; inconsistent age at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported having a birth 
before age 12, had a birth before marriage, or an interval between births of less than 
nine months. 
Finally, I restrict the sample to Hindus, who constitute about 80\% of India’s population.%
\footnote{
If the use of sex selection differ between Hindu and other groups assuming
that the baseline hazard is the same would lead to bias. The other
groups are each so small that it is not possible to estimate different
baseline hazards for each group, and are so different that combining
them into one group would not make sense.
}

In addition to the large number of women surveyed and the long period covered, a
benefit of the NFHS is that surveys enumerators pay careful attention to spacing between 
births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, is still a potentially substantial problem and
will bias both spacing and sex ratios.
As I show in the online Appendix, recall error is heavily dependent on how long ago a 
woman was married, and I, therefore, drop women married 22 years or more.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

Direct information on the use of sex selection is not available, so I compare different 
periods as a proxy, based on the changes in access and legality of prenatal sex 
determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques (PNDT) Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Although the use of sex selection increased even after 1994, some researchers argue that 
we have passed a turning point in its use \citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The allocation of spells into periods is determined by when conception, and therefore 
decisions on sex selection, can begin. 
This allocation mechanism means that some spells cover two periods, which may bias 
downward the differences between the periods.

I focus here on the second through fourth spells, that is, on the intervals from the first 
birth until at most the fourth birth.
I exclude the spell from marriage to the first birth because of data issues.
In NFHS-1 and NFHS-2 the only information on marriage is the age at marriage, and 
all first spell durations are therefore imputed.
In NFHS-3 about a third of respondent did not provide complete information on the
timing of marriage and had the length of their first spell imputed using the age at 
marriage.
The imputation results in large numbers of women with spells that are less than nine 
months and makes a comparison of spell duration over time subject to an 
excessive amount of noise.

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For all spells censoring is set at 96 months (eight years) after a birth can occur.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.

@@

@@

I divide the explanatory variables into two groups.
The first group consists of characteristics that the prior literature finds affect 
spacing choice and the use of sex selection:
mother's education, sex composition of previous children, and area of residence.
The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the age of the mother at the beginning of the spell, whether the household 
owns any land, and whether the household belongs to a scheduled tribe or caste.

>>>>

% For word counting uncomment line below and raggedright below
\usepackage[nolists]{endfloat}

@@

\date{February 2020}

@@

% Demography abstract
\noindent 
Using four rounds of India's National Family and Health Surveys and a
competing-risk hazard model, I show that average birth intervals
increased over the last four decades for all education groups. The most
significant increases are among the women most likely to use sex
selection. Despite the rise in average intervals, the likelihood of very
short spacing did not change substantially. Hence, the increases come
predominantly from the longer birth intervals getting even longer. As a
result of the longer spacing, fertility rates significantly
overestimated how fast cohort fertility fell. Although cohort fertility
and the fertility rate have started to converge, the cohort fertility is
still substantially higher than the fertility rate. Furthermore, cohort
fertility is still at or above replacement level for all but the
best-educated urban women. Finally, infant mortality risk has declined
substantially over time for all groups, but fastest for the lower
education groups, who are now close to the level of women with the most
education. Short birth spacing is still associated with higher
mortality, although the difference is small for the best-educated women.
There is no evidence that the increasing use of sex selection is
associated with higher infant mortality risk.

@@

% For word counting remove comment on the line below
\raggedright

@@

There are two primary motivations for examining birth spacing in this
situation, besides gaining a better general understanding of how
families make fertility decisions.

@@

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility behavior taking into
account both the likelihood of parity progression and the likelihood of having a child of 
a specific sex.

% [Hypotheses / questions]
There are four main sets of results.

First, birth intervals have increased for all education groups over the four decades, 
the more so, the higher the parity and the higher the education level.
Women who are most likely to use sex selection---well-educated women with no
sons---have seen the most substantial increases in birth intervals and the most 
biased sex ratios.
As a result, we now see cases that reverse the traditional spacing
pattern, with women with no sons having longer birth intervals than
women with sons.
Those least likely to use sex selection, women with less education in rural areas, 
still follow the traditional pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
Women with no sons still are the most likely to have an additional birth.
Although the probability of having a next child has declined for this group over time, the 
declines are often substantially smaller than if a boy is present.

Second, there is little difference over time and across groups in the likelihood of 
having a very short birth interval given the sex composition of prior children, 
especially for the lower parities.
Hence, the increases in average spacing come predominantly from the longer birth intervals
getting even longer over time, driven in no small degree by substantial use of sex 
selection.
The exception to this result is that the better-educated women with a \emph{very high} 
predicted use of sex selection appear to conceive later precisely because the next child 
is almost certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell 
in the 1990s and early 2000s when spacing began to increase.
Although predicted cohort fertility and the fertility rate have lately begun to converge,
the predicted cohort fertility is still ten to twenty percent higher than the
fertility rate.
Furthermore, even with the convergence, predicted cohort fertility is still at or above 
replacement level, except for the best-educated urban women.

Finally, infant mortality risk has declined substantially over time for all groups,
but fastest for the lower education groups, who are now close to the level of women with 
the most education.
The mortality risk is, however, still inversely related to education level, especially for 
very short birth intervals.
There is no evidence that the increasing use of sex selection is associated with higher 
mortality risk.

@@

The NFHS data show a U-shaped relationship between education and working for married
women, with the highest percent working for women with either no education 
or with twelve or more years and the lowest for women with eight to eleven years of 
education.%
\footnote{
See Appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}
Women are more likely to report working if they live in rural than urban areas 
and the older they are.
For married women, the percentage currently working changes relatively little over time. 
All education and age groups have, however, become substantially more likely to work for 
a family member.

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the rapid expansion in female education, and the 
associated change in the composition of better-educated women, is that the behavior
of the better-educated would change. 
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, the women adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation is an indication that this
process still operates.

@@

To provide a conceptual framework for understanding how birth spacing
may respond to the substantial changes in income, female education, and
female labor force participation that India has experienced, I discuss
in this section relevant theories and prior findings on birth spacing.

The standard economic argument for the predominant association between
increasing female education and shorter spacing is that parents incur
time costs when they have children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

The low and declining female labor force participation, especially for younger women, 
suggests, however, that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70 percent higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to their husbands' 
wages than it is positively elastic to their wages \citep{Bhargava2018}.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than the standard income effect alone predicts.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing potential income for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.
If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals 
\citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is one crucial aspect of child quality---health and mortality---where
longer spacing does lead to better outcomes, although this relationship weakens with 
maternal education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

The introduction of sex selection allows parents to avoid the birth of girls
but increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility, and, for a 
given desired number of children, the higher the parity \citep{Portner2015b}.
The increased use of sex selection in India with education and in urban
areas is consistent with a lower desired fertility for both groups
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
better enables them to access prenatal sex determination and lowers the relative costs of 
using sex selection and have long birth intervals.

We may even observe a ``delay'' effect, where parents conceive later because they know 
that short spacing is detrimental to the next child's health and that the next child is 
more likely to be a boy because of access to sex selection.
Working in the opposite direction is that women with more education can, in principle, 
space their children closer together without substantially increasing child mortality 
risk.
Given the apparent lack of pressure to return to the labor force,
this effect is, however, likely to be small.

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
As desired fertility decreases with increasing education, I, furthermore, expect 
birth spacing to increase the most among the better-educated because their household
income increases the most---even with declining female labor participation---and 
their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition does not substantially change 
this group's use of sex selection.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}
There are two problems with the standard approach in this setting.

First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell, as is assumed by the proportional hazard model.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.

The introduction of prenatal sex determination exacerbates any bias from
the proportionality assumption for two reasons. 
First, sex-selective abortion use varies across groups, which affects birth spacing. 
Second,
a household's use of sex selection may vary within a spell, which means
that the effects of covariates vary within the spell as well.

@@

The model is a discrete-time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from nine months after one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2).
\end{equation}
$D_j(t)$ is the piece-wise linear baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, and $Z_2$ captures the area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.
However, to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.%
\footnote{
I discuss the choice of which variables to use for non-proportionality in more detail
in the ``Explanatory Variables'' section.
}

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is an interval---here equal
to three months---I can estimate the model using a standard multinomial logit model.
%
% \footnote{
% The multinomial model assumes that alternative exit states are stochastically independent
% (Independence of Irrelevant Alternatives or IIA), which rules out any individual-specific 
% unmeasured or unobservable factors that affect both the hazard of having a girl and the 
% hazard of having a boy.
% I, therefore, include a proxy for fecundity discussed in Section \ref{sec:data}.
% Also, the multivariate probit model can be used as an alternative because the IIA is not 
% imposed \citep{han90}.
% The results are substantially identical between these two models and
% available upon request.
% }
In the reorganized data the outcome variable is 0 if the woman does not have a child in a 
given interval (the base outcome), 1 if she gives birth to a son in that interval, and 2 
if she gives birth to a daughter in that interval.

@@

The interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy, $b$, and of having a girl, $g$, in period $t$, conditional on 
a set of explanatory variables and not having had a child before that period.
% , as
% \begin{align}
% P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
% {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})}
% \label{eq:probability_boy} \\
% P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
% & =  
% \frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
% {1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}
% \label{eq:probability_girl}
% \end{align}
It is then straightforward to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage of boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent of births that are boys over the entire spell.

@@

The data come from the four rounds of the National Family Health Survey
collected in 1992--1993, 1998--1999, 2005--2006, and 2015--2016.
The surveys are large: 89,777, 90,303, 124,385, and 699,686 women,
respectively. NFHS-1 and NFHS-2 surveyed only ever-married women, while
the later surveys also included never-married.

I exclude visitors, as well as women in any of the following 
categories: never married; no gauna yet; married more than once; divorced; 
not living with husband; inconsistent age at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported giving birth 
before age 12, had a birth before marriage, or an interval between births of less than 
nine months. 
Finally, I restrict the sample to Hindus, who constitute about 80 percent of India’s 
population.%
\footnote{
If the use of sex selection differs across groups, assuming that the
baseline hazard is the same leads to bias. 
Furthermore, the other groups are each too small to estimate different
baseline hazards for each and so different that combining them into one
group would not make sense.
}

In addition to a large number of women surveyed and the long period covered, a
benefit of the NFHS is that enumerators pay careful attention to the spacing 
between births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, remains a potential problem.
Recall error is heavily dependent on how long ago a woman was married, and I, therefore, 
drop women married 22 years or more.%
\footnote{
See the discussion in the Appendix.
}
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

I focus on the second through fourth spells, that is, on the intervals from the first 
birth until at most the fourth birth.
I exclude the marriage to the first birth spell because many are imputed.

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place,
sterilization of the woman or her husband,
or too few births are observed for the method to work.
For all spells, censoring is set at 96 months (eight years) after a woman can first
give birth in the spell.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.

@@

Direct information on the use of sex selection is not available, so I compare different 
periods, based on the changes in access and legality of prenatal sex 
determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques (PNDT) Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Although the use of sex selection increased even after 1994, we may have passed a 
turning point in its use \citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The allocation of spells into periods is determined by when conception, and, therefore, 
decisions on sex selection, can begin. 
Hence, some spells cover two periods, which may bias downward the differences between the 
periods.

@@

I divide the explanatory variables into two groups.
The first group consists of characteristics that the prior literature finds affect 
the spacing choice and the use of sex selection:
mother's education, sex composition of previous children, and area of residence.
The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the mother's age when the spell begins, the household's
land ownership, and whether the household belongs to a scheduled tribe
or caste.
